Gravitational Wave Detection:\\ Event GW170814 and the performance of matched filtering and $\chi^2$-consistency testing in the presence of Gaussian noise, non-Gaussianly distributed noise, and transient glitches

### 2020-08-01 paper2.tex
Revised Draft

\begin{document}
\doublespacing
\title{Gravitational Wave Detection:\\
On the performance of matched filtering and $\chi^2$ consistency testing in the presence of Gaussian noise, non–Gaussianly distributed noise, and transient glitches}
\author{Vinh Nguyen}
\email[Mentored by Dr. Eric Myers – Department of Physics and Astronomy, State University of New York at New Paltz]{}
\affiliation{2020 Pioneer Academics Program \\ Klein Oak High School, Spring, TX 77379}
\date{July 15, 2020}
\begin{abstract}
Lorem ipsum dolor sit amet, consectetur adipiscing elit. Curabitur mattis quam ligula. Suspendisse tempor ante felis, nec tincidunt risus pulvinar sed. Maecenas bibendum imperdiet odio, quis cursus lacus vulputate at. Donec pharetra cursus felis eget ultrices. Phasellus semper, lorem sit amet placerat interdum, ex purus semper nunc, nec accumsan nisl purus quis tellus. Etiam ac ornare augue. Integer ullamcorper malesuada ante, at mattis dui commodo sed. Donec commodo dictum turpis ut suscipit. Pellentesque habitant morbi tristique senectus et netus et malesuada fames ac turpis egestas.
\end{abstract}
\maketitle
\section{Introduction}
Ever since Einstein's first prediction of the existence of gravitational waves in 1915, gravitational wave science and detection have been a leading scientific field of interest. With the first confirmed detection of a binary black hole merger in 2015 (event GW150914), the prospects of gravitational wave science and its applications to other fields such as astronomy and fundamental physics have only become brighter. Gravitational wave detection, however, poses many challenges. This paper explores several challenges in the detection and data analysis of gravitational waves. Specifically, it studies the performance of the optimal matched filter in different noise conditions present in the Virgo and LIGO gravitational wave detectors.

This paper uses LIGO data queried from the Gravitational Wave Open Science Center (GWOSC) is a collaborative effort between the LIGO Laboratory, the LIGO Scientific Collaboration and the Virgo Collaboration to provide the public with gravitational wave data from the two LIGO facilities (Hanford, WA and Livingston, LA) as well the Virgo Observatory in Cascina, Italy

\section{Background}
\subsection{Gravitational waves and the Laser Interferometer Gravitational-Wave Observatory (LIGO)}
In 1915, Einstein first published his theory of general relativity, describing space and time as one four-dimensional concept called \textit{``space-time''} \cite{maggiore_2008}, In this theory, he describes the force of gravity not as an inherent property that a massive body possesses, but rather as a byproduct of curved space-time. To put this more simply, we can say that a body with mass warps space-time, thereby creating a gravitational field surrounding it, affecting other bodies \cite{carlip}. To understand this idea, we can make use of an analogy: suppose that the fabric of space-time is a rubber sheet, and that any object with mass on it warps the surrounding area (FIG. 1.), hence creating a gravitational field. It follows from Einstein's Theory of Relativity that rapid changes in the curvature of space-time would produce ripples, called \textit{gravitational waves} (abbreviated as GW), that propagate outward at the speed of light  \cite{JSTORLIGO}.

@@

\begin{figure}
\includegraphics[scale = .4]{Interferometer.png}
\caption{A simplified diagram of the Michelson Fabry-Perot Interferometer. Adapted from \cite{JSTORLIGO}.}
\centering
\end{figure} 
If there is a gravitational wave passing in between the test masses, the space between them will be stretched and compressed, squeezing one arm while stretching the other such that we can calculate the value of the arm length difference $\Delta L = L_1 - L_2$ \cite{teacherintro}. To measure this quantity $\Delta L$, the device uses a technique called laser interferometry \cite{LIGOArxiv}. Specifically, a powerful laser beam is shined at the beam splitter, thereby splitting the beam in half along the two arms. These beams then each enter a Fabry-Perot cavity where they are reflected back and forth between the mirrors, building up the laser light within the interferometer and increasing the distance travelled by the beam \cite{fabry} – increasing the interferometer's sensitivity. As the laser beams in the two arms exit the Fabry-Perot cavity and return to the beam splitter, the splitter recombines the two half-beams and gives off a portion of the combined beam to a photodiode – hence allowing scientists to analyze the interference patterns of the waves using optics principles, and thus determine the quantity $\Delta L$.

@@

\subsection{LIGO Data Analysis and Conventions}
\subsubsection{The Fourier Transform}
The signal from Virgo/LIGO is recorded as a function of time. However, to process this signal, we must usually transform the signal from its original time-domain into its frequency-domain through the use of a mathematical tool called the \textit{Fourier Transform}. In simplest terms, the Fourier Transform decomposes a function into its frequency content. Given a \textit{continuous} time-series function $F(t)$, its frequency-domain representation $\widetilde{F}(f)$ is given by \cite{dsptextbook}:
\begin{equation}
\widetilde{F}(f) = \int_{-\infty}^{\infty}F(t)e^{-2\pi\iu ft} dt
\end{equation}

@@

Unfortunately, LIGO signal is not a continuous function in the time-domain, thus we have to \textit{discretize} our formulas for both the continuous forward and the continuous inverse Fourier Transforms \cite{dsptextbook}. Given a time-domain signal \textit{sequence} $f[n]$, the frequency-domain representation $\widetilde{f}[n]$ can be obtained using the discrete Fourier Transform (DFT) as follows\cite{roberts_2020}:
\begin{equation}
\widetilde{f}[n]=\sum _{n=0}^{N-1}f[n]\cdot e^{-{\frac {2\pi}{N}\iu kn}}
\end{equation}

@@

Since we are working with discrete-valued sequences, performing a DFT would give rise to spectral leakages. Thus, to minimize this, we always apply a windowing function to data in this paper whenever the use of the DFT is required (note that windowing does not decrease spectral leakage, but rather re-distributes the leakage effect so that it causes the least harm). In this paper, I also use a computer algorithm that implements the discrete Fourier Transform called the Fast Fourier Transform (FFT) which is implemented within the Python PyCBC library \cite{numpy}.

@@

In order to extract buried gravitational wave signals from noise, we must first understand the statistical properties of the noise. Noise is said to be stationary if its statistical properties do not significantly change over time. Another important noise characteristic is its Gaussianity. Given \textit{`Gaussian'} noise with a time-series function $n(t)$, both the real and imaginary components of its Fourier Transform $\widetilde{n}(f)$ follow the normal distribution \cite{yamamoto}; that is, $\Re[\widetilde{n}(f)]$ and  $\Im{[\widetilde{n}(f)]}$ have a probability density function (PDF) of:
\begin{equation}
    p(x) = \frac{1}{\sigma\sqrt{2\pi}}\exp{\left(-\frac{1}{2}\frac{(x-\mu)^2}{\sigma^2}\right)}
\end{equation}
where $\sigma^2$ is the sample variance and the mean $\mu = 0$. By the Central Limit Theorem in statistics, a set of random noise sample naturally tends to the Gaussian distribution if sufficiently large data samples were taken \cite{jaranowski2007gravitationalwave}. Nonetheless, real Virgo/LIGO noise can still deviate from normality due to abrupt noise sources such as short-duration transient noise of, sometimes, unknown origins called glitches \cite{ultimate}. Noise can also follow a variety of different probability density distributions such as the Poisson distribution and the Laplace distribution. Although it is not common in real-life Virgo/LIGO operations that these noise models are present, this paper still considers the case of Laplacian noise for theoretical analysis purposes. It is worth pointing out that over short time intervals, however, Virgo/LIGO noise can still be reasonably approximated to be Gaussian and stationary \cite{collaboration2019open}.

\subsubsection{Matched filtering}
The primary goal of Virgo/LIGO data analysis is to determine if there is gravitational wave signal buried in the noise. There are many ways that this can be accomplished; however, the most efficient and effective method is called \textit{matched filtering}, especially in cases where noise is both Gaussian and stationary \cite{helstrom_1975}. Matched filtering allows the recorded data stream to be searched for the presence of a known gravitational wave template with varying arrival time and phase offset \cite{surf}.  A matched filter requires three components: the recorded data stream $s(t)$, the \textit{power spectral density (PSD)} function $S_n(t)$ of the noise, and a known waveform template $h(t)$ of the gravitational wave. Given these quantities, the complex-valued output of the matched filter at time $t=t_0$ can be computed as follows \cite{findchirp}:
\begin{equation}
    z(t_0) = 4\int_0^\infty\frac{\widetilde{s}(f)\widetilde{h}^*(f)}{S_n(f)}e^{2\pi\iu ft_0}df
\end{equation}
where $\widetilde{h}^*(f)$ denotes the complex conjugate of the Fourier Transform of the template strain $h(t)$. Notice that the existence of $S_n(f)$ essentially down-weighs the frequencies where the noise is loud and does the opposite where the noise is weak \cite{ultimate}. Given the complex matched filter output $z(t)$, we can compute a test statistic value $\rho$ called the \textit{signal-to-noise ratio} as follows \cite{findchirp}: 
\begin{equation}
   \rho (t) = \frac{|z(t)|}{\sigma}
\end{equation}
where $\sigma$ is a normalization factor is computed from \cite{findchirp}
\begin{equation}
   \sigma^2 =  4\int_0^\infty\frac{|\widetilde{h}(f)|^2}{S_n(f)}df
\end{equation}

 Notice that here, the template waveform is computed using a set of parameters of the astrophysical source (e.g.\ the total mass, the spins, the initial frequency, etc of a binary inspiral). The goal of a matched filter search process is to see which template would give the highest SNR $\rho (t)$, and thus is mostly likely to be the strain signal in the data stream.

In the presence of non-Gaussian and/or non-stationary noise and/or glitches, the matched filter does not perform as well as expected since the filter itself is derived from Gaussian and stationary noise. A detailed mathematical derivation of the matched filter from a stationary, Gaussian noise model can be found in \textit{Brown (2007)} \cite{brown2007searching}.

There are several ways of incorporating the matched filter into a computer program that analyzes LIGO's signal. In this paper, I only make use of the matched filtering algorithm implemented in the open-source Python software package PyCBC \cite{pycbc}. This algorithm is based on the FINDCHIRP algorithm presented above and in \cite{findchirp}.

\section{Methodology}

This paper sought to evaluate the performance of the matched filtering algorithm implemented in the PyCBC software package. The evaluation focused on 3 aspects of the signal extraction output:
\begin{itemize}
    \item Performance in the presence of perfectly Gaussian noise with $\mu = 0$  and $\sigma = 1$ scaled down by a factor of $10^{-21}$.
    \item Performance in the presence of non-Gaussian noise – in this case, Laplacian noise with location $\mu = 0$ and scale parameter $b = 1,2,3$. Laplacian noise with  $\mu = 0$ and varying $b$ was chosen for the purposes of this study due to its symmetric nature (similar to Gaussian noise) and varying levels of kurtosis. The higher the scale value $b$, the heavier the Laplacian noise distribution will be at the tails (note that heavier tails correspond to higher kurtoses). These Laplace distributed noise models were all scaled down by a factor of $10^{-21}$. An illustration of these distributions are shown in FIG. 3.
    \item Performance in the presence of transient glitches. In this paper, I used five samples of known Virgo/LIGO glitches with two \textit{`Extremely Loud'} glitches, one \textit{Scattered Light} glitch, one \textit{Whistle} glitch, and one \textit{Power Line} glitch.

\end{itemize}

In the case of glitches, I employed 50 seconds around the time of peak amplitude of the glitches. A full catalog documentation of known Virgo/LIGO transient glitches was provided by Michael Zevin (PhD Candidate at Northwestern University) and the GravitySpy team \cite{gravityspy}.  The glitch samples chosen for this study are as follows:

\begin{figure}[t]
\caption{The different probability density functions $f_X(x)$ are shown below. Since $\int_{-\infty}^\infty f_X(x)dx = 1$, the lower the  functions' maximums are, the heavier their tails must be, leading to higher kurtoses.}
\includegraphics[scale = .54]{pdf graphs.png}
\centering
\end{figure}

>>>>

\doublespacing
\begin{document}

\title{Gravitational Wave Detection:\\
Event GW170814 and the performance of matched filtering and $\chi^2$-consistency testing in the presence of Gaussian noise, non-Gaussianly distributed noise, and transient glitches}
\author{Vinh Nguyen}
\email[Mentored by Dr. Eric Myers – Department of Physics and Astronomy, State University of New York at New Paltz]{}
\affiliation{2020 Pioneer Academics Program \\ Klein Oak High School, Spring, TX 77379}
\date{July 29, 2020}
\begin{abstract}
The detection of gravitational wave signals of astrophysical origins is particularly challenging because the signals are always contaminated by noise. Two of most prominent and efficient detection techniques currently used in Virgo/LIGO search pipelines are matched filtering and $\chi^2$-consistency testing. This paper investigated the performance of these two detection techniques in event GW170814 as well as in the presence of both consistently distributed noise and transient glitches. It was found that matched filtering performs best in the presence of perfect or close-perfect Gaussian noise; it can also mistake glitches for real gravitational waves if certain types of glitches are present. Additionally, it was found that $\chi^2$-consistency testing has little effects in cases where matched filtering had previously been unsuccessful. Finally, it was shown that the use of $\chi^2$-consistency testing is crucial to detecting and identifying gravitational waves in the presence of \textit{Extremely Loud} and \textit{Blip} glitches.
\end{abstract}

\maketitle
\section{Introduction}
The existence of gravitational waves was first predicted by Albert Einstein in 1916 as a direct result of his Theory of General Relativity. Since then, gravitational wave science and detection have been a leading scientific field of interest for not just physicists, but also astronomers, geologists, material scientists, and engineers as well. With the first confirmed detection of gravitational waves from a binary black hole merger in 2015 (event GW150914), the prospects of gravitational wave science and its applications to other fields have only become brighter. As of July 2020, there exists four primary fully functioning gravitational wave detectors – the LIGO Hanford Detector in Washington, the LIGO Livingston Detector in Louisiana, the Virgo Observatory in Italy, and the KAGRA detector in Japan.

The detection of gravitational waves, however, poses many challenges, for their signals are extremely weak. Gravitational wave signals can be easily contaminated by noise, thereby requiring a wide range of sophisticated mathematical and data analysis techniques to be identified. This paper explores several complications in the detection of gravitational waves – specifically, it studies the detection of event GW170814 as well as the general performance of matched filtering and $\chi^2$-consistency testing in different unfavorable noise conditions present in the Virgo and LIGO gravitational wave detectors.

This paper uses real Virgo/LIGO data queried from the Gravitational Wave Open Science Center (GWOSC). GWOSC is a collaborative effort between the LIGO Laboratory, the LIGO Scientific Collaboration and the Virgo Collaboration to provide the public with gravitational wave data from the two LIGO facilities (Hanford, WA and Livingston, LA) as well the Virgo Observatory in Cascina, Italy. The structure of this paper is as follows. \textit{Section \ref{background}} provides detailed background information necessary for the methods undertaken in this paper. Specifically, \textit{sub-section \ref{gw and ligo}} discusses gravitational waves, its detection and current detector technology while \textit{sub-section \ref{data anal}} addresses data analysis details such as  the Fourier transform, noise characteristics, matched filtering and $\chi^2$-consistency testing. \textit{Section \ref{method}} gives a detailed description of this paper's methodology and the steps undertaken. \textit{Section \ref{results}} provides the paper's preliminary results in addition to data analysis. \textit{Section \ref{conclusions}} presents the study's conclusions and proposals for possible future research directions. Finally, \textit{section \ref{summary}} gives a brief overall summary of this paper's content.

\section{Background} \label{background}
\subsection{Gravitational waves and the Laser Interferometer Gravitational-Wave Observatory (LIGO)} \label{gw and ligo}
In 1915, Albert Einstein first published his theory of general relativity, describing space and time as one four-dimensional concept called \textit{``space-time''} \cite{maggiore_2008}, In this theory, he describes the force of gravity not as an inherent property that a massive body possesses, but rather as a byproduct of curved space-time. To put this more simply, we can say that a body with mass warps space-time, thereby creating a gravitational field surrounding it, affecting other bodies \cite{carlip}. To understand this idea, we can make use of an analogy: suppose that the fabric of space-time is a rubber sheet, and that any object with mass on it warps the surrounding area (FIG. 1.), hence creating a gravitational field. It follows from Einstein's Theory of Relativity that rapid changes in the curvature of space-time would produce ripples, called \textit{gravitational waves} (abbreviated as GW), that propagate outward at the speed of light  \cite{JSTORLIGO}.

@@

\begin{figure}
\includegraphics[width = .6\textwidth]{Interferometer.png}
\caption{A simplified diagram of the Michelson Fabry-Perot Interferometer. Adapted from \cite{JSTORLIGO}.}
\centering
\end{figure} 
\begin{figure}[!tbp]
  \centering
  \begin{minipage}[b]{0.48\textwidth}
    \includegraphics[width = \textwidth]{pdf graphs.png}
    \caption{The different probability density functions $f_X(x)$ are shown. Since $\int_{-\infty}^\infty f_X(x)dx = 1$, the lower the  functions' maximums are, the heavier their tails must be, leading to higher kurtoses.}
  \end{minipage}
  \hfill
  \begin{minipage}[b]{0.48\textwidth}
    \includegraphics[width = \textwidth]{template.png}
    \caption{This figure shows the template waveform strain time-series of a binary black hole merger with masses 30 $M$\textsubscript{\(\odot\)} and 23 $M$\textsubscript{\(\odot\)} generated with PyCBC \cite{pycbc}.}
  \end{minipage}
\end{figure}
If there is a gravitational wave passing in between the test masses, the space between them will be stretched and compressed, squeezing one arm while stretching the other such that we can calculate the value of the arm length difference $\Delta L = L_1 - L_2$ \cite{teacherintro}. To measure this quantity $\Delta L$, the device uses a technique called laser interferometry \cite{LIGOArxiv}. Specifically, a powerful laser beam is shined at the beam splitter, thereby splitting the beam in half along the two arms. These beams then each enter a Fabry-Perot cavity where they are reflected back and forth between the mirrors, building up the laser light within the interferometer and increasing the distance travelled by the beam \cite{fabry} – thereby increasing the interferometer's sensitivity. As the laser beams in the two arms exit the Fabry-Perot cavity and return to the beam splitter, the splitter recombines the two half-beams and gives off a portion of the combined beam to a photodiode – hence allowing scientists to analyze the interference patterns of the waves using optics principles, and thus determine the quantity $\Delta L$.

@@

\subsection{LIGO Data Analysis and Conventions} \label{data anal}
\subsubsection{The Fourier Transform}
The signal from Virgo/LIGO is recorded as a function of time. However, to process this signal, we must usually transform the signal from its original time-domain into its frequency-domain through the use of a mathematical tool called the \textit{Fourier Transform}. In simplest terms, the Fourier Transform decomposes a function into its frequency content. Given a \textit{continuous} time-series function $F(t)$, its frequency-domain representation $\widetilde{F}(f)$ is given by \cite{dsptextbook}:
\begin{equation}
\widetilde{F}(f) = \int_{-\infty}^{\infty}F(t)e^{-2\pi\iu ft} dt
\end{equation}

@@

Unfortunately, LIGO signal is not a continuous function in the time-domain, thus we have to \textit{discretize} our formulas for both the continuous forward and the continuous inverse Fourier Transforms \cite{dsptextbook}. Given a time-domain signal \textit{sequence} $f[n]$, the frequency-domain representation $\widetilde{f}[n]$ can be obtained using the Discrete Fourier Transform (DFT) as follows\cite{roberts_2020}:
\begin{equation}
\widetilde{f}[n]=\sum _{n=0}^{N-1}f[n]\cdot e^{-{\frac {2\pi}{N}\iu kn}}
\end{equation}

@@

Since we are working with discrete-valued sequences, performing a DFT would give rise to what we call \textit{spectral leakages}. When the Fourier Transform is applied onto a finite, non-periodic data segment (i.e.\ Virgo/LIGO data), false frequency amplitudes might occur. The reason for this is that since the transformation is discrete in nature, frequencies that lies in between or across multiple frequency intervals can falsely create high amplitudes. Thus, to minimize this, we always apply a windowing function to data – tapering the data segment's ends, making it periodic – whenever the use of the DFT is required (note that windowing does not decrease spectral leakage, but rather re-distributes the leakage effect so that it causes the least harm). In this paper, I used a computer algorithm that implements the Discrete Fourier Transform called the Fast Fourier Transform (FFT) which is included within the Python PyCBC library \cite{numpy}.

\begin{figure}[!tbp]
  \centering
  \begin{minipage}[b]{0.48\textwidth}
    \includegraphics[width=\textwidth]{loud 69 graphics.png}
    \caption{The top figure shows the strain time-series of 50 seconds around Glitch 1. The bottom figure shows a Q-transformed spectrogram around Glitch 1. (Data queried from \cite{collaboration2019open})}

  \end{minipage}
  \hfill
  \begin{minipage}[b]{0.48\textwidth}
    \includegraphics[width=\textwidth]{loud 02 graphics.png}
    \caption{The top figure shows the strain time-series of 50 seconds around Glitch 2. The bottom figure shows a Q-transformed spectrogram around Glitch 2. (Data queried from \cite{collaboration2019open})}
  \end{minipage}
\end{figure}

 \begin{figure}[!tbp]
  \centering
  \begin{minipage}[b]{0.48\textwidth}
\includegraphics[width=\textwidth]{Scattered Light Graphics.png}
\caption{The top figure shows the strain time-series of 50 seconds around Glitch 3. The bottom figure shows a Q-transformed spectrogram around Glitch 3. (Data queried from \cite{collaboration2019open})}
  \end{minipage}
  \hfill
  \begin{minipage}[b]{0.48\textwidth}
\includegraphics[width=\textwidth]{power line graphics.png}
\caption{The top figure shows the strain time-series of 50 seconds around Glitch 4. The bottom figure shows a Q-transformed spectrogram around Glitch 4. (Data queried from \cite{collaboration2019open})}
  \end{minipage}
\end{figure}

 \begin{figure}[!tbp]
  \centering
  \begin{minipage}[b]{0.48\textwidth}
\includegraphics[width=\textwidth]{whistle graphics.png}
\caption{The top figure shows the strain time-series of 50 seconds around Glitch 5. The bottom figure shows a Q-transformed spectrogram around Glitch 5. (Data queried from \cite{collaboration2019open})}
  \end{minipage}
  \hfill
  \begin{minipage}[b]{0.48\textwidth}
\includegraphics[width=\textwidth]{blipppppp.png}
\caption{The top figure shows the strain time-series of 10 seconds around the Blip glitch near event GW170814 found at LIGO Livingston. The bottom figure shows a Q-transformed spectrogram around the same glitch. \cite{collaboration2019open}}
  \end{minipage}
\end{figure}

@@

In order to extract buried gravitational wave signals from noise, we must first understand the statistical properties of the noise. Noise is said to be stationary if its statistical properties do not significantly change over time. Another important noise characteristic is its Gaussianity. Given \textit{`Gaussian'} noise with a time-series function $n(t)$, both the real and imaginary components of its Fourier Transform $\widetilde{n}(f)$ follow the normal distribution \cite{yamamoto}; that is, $\Re[\widetilde{n}(f)]$ and  $\Im{[\widetilde{n}(f)]}$ have a probability density function (PDF) of:
\begin{equation}
    p(x) = \frac{1}{\sigma\sqrt{2\pi}}\exp{\left(-\frac{1}{2}\frac{(x-\mu)^2}{\sigma^2}\right)}
\end{equation}
where $\sigma^2$ is the sample variance and the mean $\mu = 0$. By the Central Limit Theorem in statistics, a set of random noise sample naturally tends to the Gaussian distribution if sufficiently large data samples were taken \cite{jaranowski2007gravitationalwave}. Nonetheless, real Virgo/LIGO noise can still deviate from this theoretical normality. Aside from the Gaussian distribution, noise can also follow a variety of different probability density distributions such as the Poisson distribution or the Laplace distribution. Although it is not common in real-life Virgo/LIGO operations that these noise models are present, this paper still considers the case of Laplacian noise for theoretical analysis purposes. Furthermore, there also exists abrupt, transient noise sources of, sometimes, unknown origins called glitches \cite{ultimate}. Glitches can be grouped into a variety of classifications according Gravity Spy's \textit{`Field Guide'} \cite{gravityspy}. In this paper, we only primarily discuss five types of glitches: \textit{Extremely Loud}, \textit{Scattered Light}, \textit{Power Line}, \textit{Whistle}, and \textit{Blip}. \textit{Extremely Loud} is a broad classification for major detector disturbances such as photodiode saturation, causing loud and high energy transient noise bursts. \textit{Scattered Light} glitches are caused by malfunctions in the optical mechanisms of Virgo/LIGO; they look like upward humps and are particularly harmful in searching for binary inspirals \cite{gravityspy}. The third classification of glitches is called \textit{Power Line}; \textit{Power Line} glitches are caused by electrical equipments such as air compressors that use alternating current (AC). The fourth classification is called \textit{Whistle} glitches; they are usually reverse V-shaped and are caused by radio signals \cite{gravityspy}. Finally, \textit{Blip} glitches are very short duration ($\approx 40$ms) noise bursts at frequencies between 30 and 500 Hz. As of currently, there are no known causes and thus \textit{Blip} glitches are the most damaging to the detection of gravitational waves \cite{gravityspy}. It is worth pointing out that over short time intervals, though, Virgo/LIGO noise can still be reasonably approximated to be Gaussian and stationary \cite{collaboration2019open}.

\begin{widetext}
\begin{figure}[t]
\includegraphics[width = .9\textwidth]{gaussian -21 template 1.png}
\caption{This figure shows the resulting cropped SNR $\rho(t)$ after matched filtering was applied to recover the template signal from perfectly Gaussian noise scaled down by a factor of $10^{-21}$. A signal peak was detected at time $t = 35.0$s with SNR $\rho = 20.0$}
\centering
\end{figure}

\begin{figure}[t]
\includegraphics[width = .9\textwidth]{chi2 gaussian zoomed.png}
\caption{This figure shows the resulting $\chi^2_r$ statistic for perfectly Gaussian noise scaled down by a factor of $10^{-21}$. A strong dip toward unity is clearly seen at the expected time $t=35s$.}
\centering
\end{figure}

\begin{figure}[t]
\includegraphics[width = .9\textwidth]{chi2 gaussian.png}
\caption{This figure shows the resulting $\chi^2_r$ statistic for perfectly Gaussian noise scaled down by a factor of $10^{-21}$, zoomed in to just the two-second interval between the time of expected signal.}
\centering
\end{figure}

\begin{figure}[t]
\includegraphics[width = .9\textwidth]{laplacian b=1 template 1.png}
\caption{This figure shows the resulting cropped SNR $\rho(t)$ after matched filtering was applied to recover the template signal from perfectly Laplacian noise with $b=1$, scaled down by a factor of $10^{-21}$. A signal peak was detected at time $t = 35.0$s with SNR $\rho = 12.6$.}

\centering
\end{figure}

\begin{figure}[t]
\includegraphics[width = .9\textwidth]{chi2 laplace b=1 zoomed.png}
\caption{This figure shows the resulting $\chi^2_r$ statistic for perfectly Laplacian noise with $b=1$ scaled down by a factor of $10^{-21}$. A strong dip toward unity is seen at time $t = 35.0$s with climbing values just around it.}
\centering
\end{figure} 
\end{widetext}

\begin{widetext}

\begin{figure}[t]
\includegraphics[width = .9\textwidth]{chi2 laplace b=1.png}
\caption{This figure shows the resulting $\chi^2_r$ statistic for perfectly Laplacian noise with $b=1$ scaled down by a factor of $10^{-21}$. The graph is zoomed in to show just the two-second time interview around the time of expected signal. A strong dip toward unity is seen at time $t = 35.0$s.}
\centering
\end{figure}

\begin{figure}[t]
\includegraphics[width = .9\textwidth]{laplacian b=2 template 1.png}
\caption{This figure shows the resulting cropped SNR $\rho(t)$ after matched filtering was applied to recover the template signal from perfectly Laplacian noise with $b=2$, scaled down by a factor of $10^{-21}$. A signal peak was detected at time $t = 35.0$s with SNR $\rho = 6.1$.}
\centering
\end{figure}

\begin{figure}[t]
\includegraphics[width = .9\textwidth]{chi2 laplace b=2 zoomed.png}
\caption{This figure shows the resulting $\chi^2_r$ statistic for perfectly Laplacian noise with $b=2$ scaled down by a factor of $10^{-21}$. A small dip toward unity is seen at time $t = 35.0$s.}
\centering
\end{figure} 
\end{widetext}
\subsubsection{Matched filtering}
The primary goal of Virgo/LIGO data analysis is to determine if there is gravitational wave signal buried in the noise. There are many ways that this can be accomplished; however, the most efficient and effective method is called \textit{matched filtering}, especially in cases where noise is both Gaussian and stationary \cite{helstrom_1975}. Matched filtering allows the recorded data stream to be searched for the presence of a known gravitational wave template with varying arrival time and phase offset \cite{surf}. Matched filtering requires three components: the recorded data stream $s(t)$, the \textit{power spectral density (PSD)} function $S_n(t)$ of the noise, and a known waveform template $h(t)$ of the gravitational wave. (Note that the PSD is essentially a measure of how power is distributed at different frequencies of the signal; detailed mathematical descriptions can be found in \cite{psd}.) Given these quantities, the complex-valued output of matched filtering at time $t=t_0$ can be computed as follows \cite{findchirp}:
\begin{equation}
    z(t_0) = 4\int_0^\infty\frac{\widetilde{s}(f)\widetilde{h}^*(f)}{S_n(f)}e^{2\pi\iu ft_0}df
\end{equation}
where $\widetilde{h}^*(f)$ denotes the complex conjugate of the Fourier Transform of the template strain $h(t)$. Notice that the existence of $S_n(f)$ in the denominator essentially down-weighs the frequencies where the noise is loud and does the opposite where the noise is weak \cite{ultimate}. Given the complex matched filter output $z(t)$, we can compute a test statistic value $\rho$ called the \textit{signal-to-noise ratio} as follows \cite{findchirp}: 
\begin{equation}
   \rho (t) = \frac{|z(t)|}{\sigma}
\end{equation}
where $\sigma$ is a normalization factor is computed from \cite{findchirp}
\begin{equation}
   \sigma^2 =  4\int_0^\infty\frac{|\widetilde{h}(f)|^2}{S_n(f)}df
\end{equation}

 Notice that here, the template waveform is computed using a set of parameters of the astrophysical source (e.g.\ the total mass, the spins, the initial frequency, etc of a binary inspiral). The goal of the matched filtering search process is to see which template would give the highest SNR $\rho (t)$, and thus is mostly likely to be the strain signal in the data stream.

In the presence of non-Gaussian and/or non-stationary noise and/or glitches, matched filtering does not perform as well as expected since the filter itself is derived from Gaussian and stationary noise. A detailed mathematical derivation of the matched filter from a stationary, Gaussian noise model can be found in \textit{Brown (2007)} \cite{brown2007searching}.

There are several ways of incorporating matched filtering into a computer program that analyzes LIGO's signal. In this paper, I only make use of the matched filtering algorithm implemented in the open-source Python software package PyCBC \cite{pycbc}. This algorithm is based on the FINDCHIRP algorithm presented above and in \cite{findchirp}.
\begin{widetext}
\begin{figure}[t]
\includegraphics[width = .9\textwidth]{chi2 laplace b=2.png}
\caption{This figure shows the resulting $\chi^2_r$ statistic for perfectly Laplacian noise with $b=2$ scaled down by a factor of $10^{-21}$. The graph is zoomed in to show just the two-second time interview around the time of expected signal. A relatively weak dip toward unity is seen at time $t = 35.0$s.}
\centering
\end{figure}

\begin{figure}[t]
\includegraphics[width = .9\textwidth]{laplacian b=3 template 1.png}
\caption{This figure shows the resulting cropped SNR $\rho(t)$ after matched filtering was applied to recover the template signal from perfectly Laplacian noise with $b=3$, scaled down by a factor of $10^{-21}$. A signal peak was detected at time $t = 21.7$s with SNR $\rho = 4.6$.}
\centering
\end{figure}
\end{widetext}

\subsubsection{$\chi^2$-consistency testing and SNR re-weighting} \label{chisqaured}

As described in the previous section on matched filtering, the algorithm breaks down in the presence of spurious noise sources such as glitches and non-Gaussian and/or non-stationary noise. When this happens, matched filtering gives rise to elevated signal-to-noise peaks that are due to noise instead of gravitational waves of astrophysical origins, further complicating the detection data analysis pipeline. To address this issue, many existing detection pipelines, such as the PyCBC pipeline \cite{veto}, employ the use of signal consistency testing to distinguish between real signals and false ones due to noise. There are different methods to perform consistency testing on Virgo/LIGO data. In this paper, we make use of the $\chi^2$ time-frequency discriminator test described with great details in \textit{Allen (2005)} \cite{Allen_2005} and implemented in PyCBC's search pipelines \cite{pycbc}. The general purpose of the $\chi^2$ test is to see how much different ranges of frequencies of the template waveform contribute to the signal-to-noise ratio $\rho(t)$. The overall algorithm goes as follows. We first divide our template up into $p$ frequency sub-intervals (or\textit{ frequency bins}) $\Delta f_1,\Delta f_2,..., \Delta f_p$ with corresponding signal-to-noise ratios $\rho_1,\rho_2,...,\rho_p$. Then, we compute the extent to which a particular frequency bin contributes to $\rho(t)$ by calculating a $\chi^2$ test statistic as follows:
\begin{equation} 
    \chi^2 = \sum^p_{i=0}\left(\left|\rho_i - \frac{\rho}{p}\right|\right)^2
\end{equation}
where $\chi^2$ has a classical $\chi^2$ probability density distribution with degrees of freedom $\nu = 2(p-1)$ since the signal-to-noise ratio $\rho$ is a complex valued number. If the template matches closely with signal-to-noise output at a certain bin, the $\chi^2$ test statistic should be close to $\nu$. In this paper, we will be normalizing $\chi^2$ as follows 
\begin{equation}
    \chi^2_{r} = \frac{\chi^2}{\nu}
\end{equation}

The reduced chi-squared value $\chi^2_r$ would then approach unity when the template matches well with the signal-to-noise and deviate from that when it does not.

Using the reduced $\chi_r^2$ value, we can then `re-weigh' our signal-to-noise ratio $\rho(t)$ by utilizing the fact that $\chi_r^2$ describes the degree to which the SNR deviates from the template. The `re-weighted' SNR, denoted $\hat{\rho}$, is computed by scaling $\rho$ as follows:
\begin{equation}
    \hat{\rho} = \rho \,\cdot \,
    \begin{cases}
    1     & \quad\quad\text{if}\quad \chi^2_r\leq1\\
     \left[\frac{1}{2} + \frac{1}{2}(\chi^2_r)^3\right]^{-1/6}   & \quad\quad\text{if}\quad \chi^2_r>1
    \end{cases}
\end{equation}

Note that this process of re-weighing the SNR down-weighs $\rho$ where it is not strongly matched by the template used in the computation of $\rho$ and does the opposite where the template closely aligns with $\rho(t)$.

\begin{figure}[t]
\includegraphics[width = .9\textwidth]{chi2 laplace b=3 zoomed.png}
\caption{This figure shows the resulting $\chi^2_r$ statistic for perfectly Laplacian noise with $b=3$ scaled down by a factor of $10^{-21}$. A very weak dip toward unity is seen at time $t = 35.0$s.}
\centering
\end{figure}

\begin{figure}[t]
\includegraphics[width = .9\textwidth]{chi2 laplace b=3.png}
\caption{This figure shows the resulting $\chi^2_r$ statistic for perfectly Laplacian noise with $b=3$ scaled down by a factor of $10^{-21}$. The graph is zoomed in to show just the two-second time interview around the time of expected signal. A tiny dip toward unity is seen at time $t = 35.0$s.}
\centering
\end{figure}

\begin{figure}[t]
\includegraphics[width = .9\textwidth]{glitch loud 69 template 1.png}
\caption{This figure shows the resulting cropped SNR $\rho(t)$ after matched filtering was applied to recover the template signal from Glitch 1. A signal peak was detected at time $t = 17.7$s with SNR $\rho = 305.0$.}
\centering
\end{figure}

\begin{figure}[t]
\includegraphics[width = .9\textwidth]{glitch loud 02 template 1.png}
\caption{This figure shows the resulting cropped SNR $\rho(t)$ after matched filtering was applied to recover the template signal from Glitch 2. A signal peak was detected at time $t = 17.7$s with SNR $\rho = 294.6$.}
\centering
\end{figure}
